---
title: "EDAV Project 1"
output: html_document
---

```{r}
#############################
# < Waitlist Survivors >
#
# STAT W4701 
# Homework < 01 >
# < Feb 11th >
#
#############################
#setwd("~/Documents/Study/Spring 2016/EDAV")

library(xlsx)
library(reshape)


# Read xlsx file
raw.data <- read.xlsx("Survey+Response.xlsx", sheetName = "Form Responses 1")
data <- raw.data


# drop empty columns
data <- data[, colSums(is.na(data)) < nrow(data)]


# Clean up variable names
cleaned.names <- c('waitlist', 'program','tools','exp.R.manipulation',
          'gender','text.editor','exp.R.graphics','exp.R.advanced','exp.R.reproducible','exp.Matlab','exp.Github')
names(data) <- cleaned.names

library(stringr)
# Split tools column into multiple columns
## Count the maximum number of commas per row in the tools column
num.commas <- max(sapply(gregexpr(",", data$tools, fixed = TRUE), function(x) max(length(x)))) + 1
## Split the tools column in separate columns
tool.cols <- colsplit(data$tools,"[,]",names=sapply(1:num.commas, function(x) paste('tools', x, sep='')))
tool.cols <- sapply(tool.cols, str_trim)
unique.tool.values <- unique(unlist(lapply(tool.cols, FUN=unique)))

tool.matrix <- matrix(0, nrow = nrow(tool.cols),20)

tool.names <- unique.tool.values
colnames(tool.matrix) <- tool.names
tool.matrix <- as.data.frame(tool.matrix)
for (i in 1:ncol(tool.cols) ) {
      for (tool.name in tool.names) {
          tool.matrix[which(tool.cols[, i] == tool.name), tool.name] <- 1
          }
}

clean.tool.names <- c("R", "Excel", "Matlab", "RStudio", "Github", "SQL", "Shell", "Python", "SPSS", "ggplot2", "GDrive", "Web", "C", "Dropbox", "Regex", "Stata", "LaTeX", "Sweave", "XML", "Lattice")
names(tool.matrix) <- clean.tool.names
data <- cbind(data, tool.matrix)
data <- subset(data, select=-c(tools))


# Clean up inconsistent program values
levels(data$program)
sort(table(data$program))
levels(data$program)[levels(data$program)=="Ms in ds"] <- "IDSE (master)"
levels(data$program)[levels(data$program)=="MSDS"] <- "IDSE (master)"
levels(data$program)[levels(data$program)=="PhD Biomedical Informatics"] <- "Ph.D."
levels(data$program)[levels(data$program)=="Data Science"] <- "Unknown"


# Clean up text editor values
levels(data$text.editor)
sort(table(data$text.editor))
# Remove extra spaces
data$text.editor <- factor(str_trim(data$text.editor))

levels(data$text.editor)[levels(data$text.editor)=="textwrangler"] <- "TextWrangler"
levels(data$text.editor)[levels(data$text.editor)=="Text Wrangler"] <- "TextWrangler"
levels(data$text.editor)[levels(data$text.editor)=="textWrangler"] <- "TextWrangler"

levels(data$text.editor)[levels(data$text.editor)=="Sublime"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="Sublime Text 2"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="sublime"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="Sublime Text!"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="sublime text 2"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="Sublime 2"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="sublime text"] <- "Sublime Text"

levels(data$text.editor)[levels(data$text.editor)=="I used jupyter last semester"] <- "Jupyter"
# Ipython is now known as Jupyter
levels(data$text.editor)[levels(data$text.editor)=="Ipython"] <- "Jupyter"

levels(data$text.editor)[levels(data$text.editor)=="haven't used any"] <- "None"






# Assign level ordering to experience variables
experience.cols <- names(data)[grepl("^exp.", names(data))]
for (exp.col in experience.cols) {
  data[, exp.col] <- factor(data[, exp.col], levels=levels(data[, exp.col])[c(4, 1, 2, 3)], ordered=TRUE)
}

```


# Chris

```
barplot(table(data$waitlist))
barplot(table(data$exp.R.manipulation))
barplot(table(data$exp.R.advanced, data$program))


barplot(sort(sapply(data[11:30], sum), decreasing=FALSE), horiz=TRUE)

# Correlation plot
library(ellipse)
ctab <- cor(data[, 11:30])
distance <- as.dist(1 - ctab)
# http://datascienceplus.com/hierarchical-clustering-in-r/
plot(hclust(distance), 
     main="Dissimilarity = 1 - Correlation", xlab="")
ordering <- hclust(distance)$labels[hclust(distance)$order]

round(ctab, 2)
colorfun <- colorRamp(c("#CC0000","white","#3366CC"), space="Lab")
plotcorr(ctab[ordering,ordering], col=rgb(colorfun((ctab[ordering,ordering]+1)/2), maxColorValue=255),
         mar = c(0.1, 0.1, 0.1, 0.1))




# PCA
tools.pca <- prcomp(data[, 11:30],
                 center = TRUE,
                 scale. = TRUE)

# Scree plot
plot(tools.pca)

# Cumluative proportion of variance explained. Let's go with 2
summary(tools.pca)

print(tools.pca)



tools.pca.rotation <- data.frame(tools.pca$rotation[, 1:6])
tools.pca.rotation['label'] <- row.names(tools.pca.rotation)
tools.pca.rotation.molten <- melt(tools.pca.rotation, id.vars='label')


ggplot(tools.pca.rotation.molten, aes(x=label, y=value)) + 
    geom_bar(position='identity', stat = "identity", alpha=.3) +
  coord_flip() + 
  facet_grid(. ~ variable)
  

table(data$program, data$SPSS)


library(devtools)
install_github("ggbiplot", "vqv")
library(ggbiplot)


tools.pca.x <- data.frame(tools.pca$x)[, 1:6]
tools.pca.x['label'] <- data$program
tools.pca.x.molten <- melt(tools.pca.x)

ggplot(tools.pca.x.molten, aes(value, color=label)) +
    #geom_bar(position='identity', stat = "identity", alpha=.3) +
  geom_density() +
  coord_flip() +
  facet_grid(. ~ variable)




g <- ggbiplot(tools.pca, choices = c(2, 3), scale=1, var.scale = 1, 
              groups = data$program, ellipse = TRUE, 
              circle = TRUE)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
               
```

# Jose

```{r results="asis" }
#install.packages("rjson")
#install.packages("rCharts")
#install.packages("devtools")
#install.packages("Rcpp")
#install_github('ramnathv/rCharts')


library(devtools)
library(Rcpp)
require('httr')
require(rCharts)
require(rjson)
# Donut with Programs
pie.program <- rCharts$new()
pie.program <- nPlot(~program, data = data, type = 'pieChart')
pie.program$chart(donut = TRUE)
pie.program$print('iframe',  include_assets=TRUE)
```
Donut chart explains one of the most important features in our dataset. Almost 50% of our students are Master in data-science, and 20% are in certificate program.
```{r results="asis" }
#Sankey Graph
data_t=as.data.frame(table(data$program,data$text.editor))
colnames(data_t) <- c("source", "target", "value")
data_t=data_t[data_t[,3]>0,]
#remove all data with 0 relations
sankeyPlot <- rCharts$new()
sankeyPlot$setTemplate(script = "./rCharts_d3_sankey-gh-pages/layouts/chart.html")
sankeyPlot$setLib('http://timelyportfolio.github.io/rCharts_d3_sankey/libraries/widgets/d3_sankey')
sankeyPlot$set(
  data = data_t,
  nodeWidth = 15,
  nodePadding = 10,
  layout = 32,
  width = 750,
  height = 500,
  units = "TWh",
  title = "Editor-Program"
)

sankeyPlot$save('sankey.html',  cdn = TRUE)





```
<iframe width="900" height="520" frameborder="0" src = 'sankey.html' ></iframe>
The sankey chart is a good way to present a relationship between two o more variables, in this particular case  *program* and *text Editor*. it shows a preference by RStudio that is explained by the huge number of statistic and data-science master. However few students of statistics prefer different editors compare with Data-science where 40% of the students do not use RStudio.
``````{r results="asis" }
col.names.data=c("exp.R.advanced","exp.Github","exp.R.reproducible"
                 ,"exp.R.manipulation","exp.R.graphics","exp.Matlab")
matrix={}
for (name.col in col.names.data){
  row=cbind(name.col,cbind(as.character(data[,"program"]),as.character(data[,name.col])))
  matrix=rbind(matrix,row)
}
data.splot.tool=as.data.frame(matrix)
colnames(data.splot.tool)=c("tool","program","use")
data.splot.f=data.splot.tool[data.splot.tool$use != "None",]
data.final=as.data.frame(table(data.splot.f$program,data.splot.f$tool))
data.s=as.data.frame(table(data.splot.tool$tool,data.splot.tool$use))
colnames(data.s) <- c("source", "target", "value")
colnames(data.final) <- c("source", "target", "value")
data_sfinal=rbind(data.s,data.final)

#remove all data with 0 relations
sankeyPlotTool <- rCharts$new()
sankeyPlotTool$setTemplate(script = "./rCharts_d3_sankey-gh-pages/layouts/chart.html")
sankeyPlotTool$setLib('http://timelyportfolio.github.io/rCharts_d3_sankey/libraries/widgets/d3_sankey')
sankeyPlotTool$set(
  data = data_sfinal,
  nodeWidth = 15,
  nodePadding = 10,
  layout = 35,
  width = 750,
  height = 500,
  units = "TWh",
  title = "Editor-Program"
)

sankeyPlotTool$save('sankey_tool.html',  standalone = TRUE)


```
<iframe width="900" height="520" frameborder="0" src = 'sankey_tool.html' ></iframe>
This Chart shows a relation between *program* , *tool* and *experience level*. We removed the association between program and tool when the level of experience is None. In Master Data-Science (IDSE) Github and Matlab are the most unknown topic and in certification program have a poor experience with R.reproducible. However 30% of all students do not have any knowledge in Matlab and 91% of them have worked with R.Manipulation.
``````{r results="asis" }



col.names.data=c("exp.R.advanced","exp.Github","exp.R.reproducible"
                 ,"exp.R.manipulation","exp.R.graphics","exp.Matlab")
matrix={}
for (name.col in col.names.data){
  row=cbind(name.col,as.character(data[,name.col]))
  matrix=rbind(matrix,row)
}
data.boxplot=as.data.frame(matrix)
colnames(data.boxplot)=c("tool","experience")

n2<-rCharts$new()
n2 = nPlot( ~experience,group="tool", data = data.boxplot, type = "multiBarChart")
n2$print('inline',  include_assets=TRUE)

```
The Multibarplot is used to summarize the experience  with out taking account the program. R manipulation and graphics are topics that the students are confident. The number of confident students in any tool is similar to the number of students with a lack experience. Little is the most likely answer


# Phoebe
```
## 
par(mar=c(3,3,3,3))
barplot(table(data$waitlist))
barplot(table(data$exp.R.manipulation))
barplot(table(data$gender))
barplot(table(data$text.editor),cex.names = 0.6)
barplot(table(data$exp.R.graphics), xlab = "experience with R (graphics)")
barplot(table(data$exp.R.advanced),xlab = "experience with R (advanced)")
barplot(table(data$exp.R.reproducible),xlab = "experience with R (reproducible)")
barplot(table(data$exp.Matlab),xlab = "experience with Matlab")
barplot(table(data$exp.Github),xlab = "experience with Github")
tools = data.matrix(data)[,11:30]
barplot(tools,xlab="Tools", names.arg = as.vector(colnames(as.matrix(data)[,11:30])),
        ylab="Number of Students" , ylim = c(0,80),cex.names = 0.6, col="light blue")

## Following codes are used to create a pie chart to show the distribution of the use of tools. There are lots of tools are not common or not used in the students of these class. Therefore I combine all those into one category. 
par(mar=c(0.5,0.5,0.5,.5))
t <-colSums(tools)
pie(colSums(tools),radius = 1)
s.t <- as.matrix(sort(t,decreasing = T))
other<-sum(s.t[7:20,1])
s.t[7,1] <- other
s.t<-as.matrix(s.t[1:7,1])
rownames(s.t)<-c("R","Matlab","Excel","Github","SQL","Rstudio","other")
pie(s.t[1:7,])
## R is still the most common choice among the students in this class. Matlab is the next. Majority of the studetns don't have enought experience with the github. 

## import library "ggplot2"
library(ggplot2)
## all of the following codes are designed to illustrate the experience with each tool.
qplot(data$exp.R.graphics, data = data) ## majority of the student have limited expereince with the R.graphics.
qplot(data$waitlist) # the waitlist for this class was not very long.
qplot(data$program) # the diversity of the students is not that big. It's understandable since this is class for DS major specifically.
qplot(data$exp.R.advanced) # Similar to the R graphics skills, students have limited experiene in advanced packages in R as well.
qplot(data$exp.R.manipulation) #Contrary to the previous two categories, more students have confidence and experience with data manipulation in R.
qplot(data$gender) #the number of male students are more than twice of the girls.
qplot(data$text.editor) # the number of studetns using Rstudio are lot more than any other text editor tools.
qplot(data$exp.R.reproducible) # it seems like that this part of the R functions are less common than other three.
qplot(data$exp.Matlab) # Matlab is not common choice among the students in the past.
qplot(data$exp.Github) # Same for the Github.
```


# Yusen
```{r}
# look enrollment condition by gender.
by_gender <- as.data.frame(as.character(data$gender[which(data$gender != "doesn't matter")]))
colnames(by_gender) <- "gender"

barplot(table(by_gender), xlab = "Gender", ylab = "Total number", main = "Enrollment by gender")

# find out top 7 code editor tools used most.
by_preferred_tool <- table(data$text.editor)
by_preferred_tool <- sort(by_preferred_tool, decreasing = TRUE)[1:7]
barplot(by_preferred_tool, las = 3, cex.names = 0.74, ylab = "Total number", main = "top 7 code editor tools used most")

opar <- par()
par(mfrow = c(2,3))
# Experience with R manipulation.
r_manipulation <- table(data[ , 3])
barplot(r_manipulation, las = 3, cex.names = 0.9, ylab = "Total number", main = "Experience with R manipulation")

# Experience with R graphics. 
r_graphics <- table(data[ , 6])
barplot(r_graphics, las = 3, cex.names = 0.9, ylab = "Total number", main = "Experience with R graphics")

# Experience with R advanced.
r_advanced <- table(data[ , 7])
barplot(r_advanced, las = 3, cex.names = 0.9, ylab = "Total number", main = "Experience with R advanced")

# Experience with R reproducible.
r_reproducible <- table(data[ , 8])
barplot(r_reproducible, las = 3, cex.names = 0.9, ylab = "Total number", main = "Experience with R reproducible")

# Experience Matlab.
matlab <- table(data[ , 9])
barplot(matlab, las = 3, cex.names = 0.9, ylab = "Total number", main = "Experience with Matlab")

# Experience with Github.
github <- table(data[ , 10])
barplot(github, las = 3, cex.names = 0.9, ylab = "Total number", main = "Experience with Github")

par(opar)

# Programming and Analytical Experiences with All Tools.
library(vcd)
exp_tool <- apply(data[ , c(3, 6:10)], 2, FUN = table)
colnames(exp_tool) <- c("r_manipulation", "r_graphics", "r_advanced", "r_reproducible", "Matlab", "Github")
spine(t(exp_tool), main = "Programming and Analytical Experiences with All Tools")

# Total numer of use with all tools.
barplot(sort(colSums(data[, 11:30]), decreasing = TRUE), las = 3, cex.names = 0.9, ylab = "Total number", main = "Experience with all tools")
```




# Manu


```{r}
library(ggplot2)
library(cowplot)

# skills plot vs program of the student

#comp.R.exp <- data[,2:10]

```

```{r}
comp.R.exp <- data[,2:10]

```

```{r}
g1 <- ggplot( comp.R.exp, aes(exp.R.manipulation, fill= program))+geom_bar()+theme(legend.position= "none",text=element_text(size=10),axis.text.x= element_text(size=10))

g2 <- ggplot( comp.R.exp, aes(exp.R.graphics, fill= program))+geom_bar()+theme(legend.position= "none",text=element_text(size=10),axis.text.x= element_text(size=10))

g3 <- ggplot( comp.R.exp, aes(exp.R.advanced, fill= program))+geom_bar()+theme(legend.position= "none",text= element_text(size=10),axis.text.x= element_text(size=10))

g4 <- ggplot( comp.R.exp, aes(exp.R.reproducible, fill= program))+geom_bar()+theme(legend.position= "none",text=element_text(size=10),axis.text.x= element_text(size=10))

g5 <- ggplot( comp.R.exp, aes(exp.Matlab, fill= program))+geom_bar()+theme(legend.position= "none",text=element_text(size=10),axis.text.x= element_text(size=10))

g6 <- ggplot( comp.R.exp, aes(exp.Github, fill= program))+geom_bar()+theme(text=element_text(size=10),axis.text.x= element_text(size=10))


require(cowplot)
plot_grid(g1,g2,g3,g4,g5,g6,labels=c("Manipulation","Graphics","Advanced R","Markdown","Matlab","Github"),label_size=10, ncol=2,nrow=3)

```

```{r}
g7 <- ggplot( data, aes(exp.R.manipulation, fill= gender))+geom_density(alpha=0.5, lwd=0.8, adjust=0.5)+theme(text=element_text(size=10),axis.text.x= element_text(size=7))
g8 <- ggplot( data, aes(exp.R.advanced, fill= gender))+geom_density(alpha=0.5, lwd=0.8, adjust=0.5)+theme(text=element_text(size=10),axis.text.x= element_text(size=7))
g9 <- ggplot( data, aes(exp.Matlab, fill= gender))+geom_density(alpha=0.5, lwd=0.8, adjust=0.5)+theme(text=element_text(size=10),axis.text.x= element_text(size=7))
g10 <- ggplot( data, aes(exp.Github, fill= gender))+geom_density(alpha=0.5, lwd=0.8, adjust=0.5)+theme(text=element_text(size=10),axis.text.x= element_text(size=7))

plot_grid(g7,g8,g9,g10,labels=c("Manipulation","Advanced R","Matlab","Github"),label_size=10)


```


